# Introduction to causal mediation analysis

```{r setup}
#| include: false
#| cache: false
library(here)
library(DiagrammeR)
```

In this session we will go over fundamental concepts of causal inference
in general and then focus more specifically on causal mediation
analysis.

## What is a cause?

We employ causal inference constantly in our daily lives. Consider a
simple example: when you press a light switch, you expect the light to
turn on. Or in another scenario, you anticipate that performing action A
will result in outcome Y. In some instances, we can identify
deterministic relationships where action A will lead to outcome Y. While
these straightforward cause-and-effect relationships are intuitive and
help us navigate the world, they become significantly more complex when
we explore health outcomes.

::: callout-note
## Notation

A = received treatment/intervention/exposure (e.g., 1 = intervention, 0
= no intervention)

Y = observed outcome (e.g., 1 = developed the outcome, 0 = no outcome)

$Y^{a=1}$ = Counterfactual outcome under treatment a=1 (i.e., the
outcome had everyone, counter to the fact, received treatment a = 1)

$Y^{a=0}$ = Counterfactual outcome under treatment a =0 (i.e., the
outcome had everyone, counter to the fact, received treatment a = 0)
:::

### Individual causal effect

When investigating health outcomes, we would ideally want to know if you
**do X**, then Y will happen. We could have a specific question:

> Will eating more red meat give me higher blood glucose in 1 year?

To answer this question we would ideally have you consume more red meat
over 1 year and measure your blood glucose levels. Then, we would turn
back time, and make you eat something else over 1 year and then measure
your blood glucose levels again. If there is a difference between your
two outcomes, then we say there is a causal effect.

But we can never do this in the real world.

### Average causal effect

Instead, we can perform a randomized controlled trial. We now ask a
slightly different question:

> Will eating more red meat give adults higher blood glucose in 1 year?

We can randomely assigning one group to consume more red meat and the
other group to consume more of something else over 1 year. Then we
compare the average blood glucose levels after 1 year in each of the
groups. If there is a difference, we could say there is an average
causal effect.

::: callout-note
## Notation

We now modify the terms a little.

$E[Y^{a=1}]$ the average counterfactual outcome, had all subjects in the
population received treatment a = 1.

$Pr[Y^{a=1}]$ the proportion of subjects that would have developed the
outcome Y had all subjects in the population of interest received
treatment a = 1.

$E[Y^{a=0}]$ the average counterfactual outcome, had all subjects in the
population received no treatment a = 0.

$Pr[Y^{a=0}]$ the proportion of subjects that would have developed the
outcome Y had all subjects in the population of interest received no
treatment a = 0.
:::

### Definition of a causal effect

More formally we can now define a causal effect @hernan_definition_2004:

$E[Y^{a=1} = 1] - E[Y^{a=0} = 1] \ne 0$

## Association vs causation

What makes it complicated to estimate a causal effect is that we cannot
observe the outcome under different treatments.

When we only have a subset of the outcomes, we have an association. This
is illustrated in @fig-causation-association.

\![causation vs
association\](/images/causation-association.png){#fig-causation-association}

![Causation vs association. From
@hernan_definition_2004](/images/causation-association.PNG){#fig-causation-association}

If we want to infer a causal effect (i.e., what would have happened, had
everyone done A=1 vs A=0), we need three assumptions to be fulfilled:

-   Exchangeability

-   Consistency

-   Positivity

    ::: callout-note
    ## Causal Assumptions

    -   Exchangeability

    The risk of the outcome in A = 1 would have been the same as the
    risk of the outcome in A = 0, had those in A = 1 received A = 0.

    Think about a randomized trial where you, by mistake, give the
    intervention to the other group. The effect should be the same as
    the one you would have observed had the groups been correct.

    We can also have conditional exchangeability. The risk is similar in
    subsets of the population. That could be within the levels of a
    variable W (e.g., education).

    -   Consistency

    The treatments under comparison are well-defined and correspond to
    the versions of treatment observed in the data:

    -   Precise definition of $Y^a$ via a
    -   Link counterfactual outcomes with observed outcomes

    $Y^a = Y$ for every individual with A = a.

    The observed outcome for all treated equals the outcome if they had
    received the treatment.

    -   Positivity

    The probability of receiving every value of treatment conditional on
    L is greater than zero. In other words, there must be a probability
    of being assigned to each treatment level.

    Image if we don't have anyone with A = 0 among those with L=1 (e.g.
    long education), then we cannot estimate the conditional probability
    of the outcome.
    :::

## Causal mediation analysis

::: callout-note
## recall the major limations of tranditional mediation methods:

-   Non-linearity
-   Interactions
-   Confounding
-   Multiple mediators
:::

Causal mediation analysis is an extension of the traditional approach
by:

::: callout-note
-   Outlining all confounding assumptions needed
-   Handling non-linearity and interaction
-   Clearly defining estimands of interest
:::

## Confounding assumptions

By using DAGs, the assumptions about confounding are made much more
explicit. DAG under no mediator-outcome relation affected by treatment:

```{r}
#| echo: false

grViz("
digraph {
  graph []
  node [shape = plaintext]
    W [label = 'W']
    A [label = 'A']
    M [label = 'M']
    Y [label = 'Y']
  edge [minlen = 2]
    A->M
    M->Y
    A->Y
    W->A
    W->Y
    W->M
{rank = same; W; A; M; Y; }
}
")
```

We can see that we not only have to take confounding between the
treatment and outcome into account, but we also have to take
mediator-outcome ($A \leftarrow W \rightarrow Y$) confounding into
account.

In addition, we can have a more complicated situation where the
treatment also impacts another mediator that is also a mediator-outcome
confounder.

Below DAG shows where the mediator-outcome relation is affected by
treatment:

```{r}
#| echo: false

grViz("
digraph {
  graph []
  node [shape = plaintext]
    W1 [label = 'W1']
    A [label = 'A']
    M [label = 'M']
    Y [label = 'Y']
    W2 [label = 'W2']
  edge [minlen = 2]
    A->M
    M->Y
    A->Y
    W1->A
    W1->Y
    W1->M
    W2->M
    W2->Y
    A->W2
    W1->W2
{rank = same; W1; A; M; Y; }
{rank = max; W2 ; }
}
")
```

From the DAG rules, we have a special problem that we cannot solve with
traditional regression approaches. If we adjust for W2 we open a
backdoor path by adjusting for the collider
$W1 \rightarrow W2 \leftarrow A$. We will work on how to solve this
problem later in the course.

## Non-linearity and interactions

Neither the product method nor the difference method in traditional
mediation analysis can effectively account for interaction effects or
non-linear relationships between variables.

Causal mediation analysis addresses these limitations through more
sophisticated approaches. Specifically, it can:

-   Utilize regression-based methods and alternative causal inference
    techniques like g-computation
-   Fundamentally differ from traditional approaches by:
    -   Constructing a comprehensive causal model that explicitly
        incorporates non-linear relationships and interaction effects
    -   Artificially manipulating the data to set the treatment and the
        mediator to certain values
    -   Predicting outcomes using the developed causal model
    -   Comparing predicted outcomes to understand causal mechanisms

Note: additional approaches also exists, but we will not focus on these
in this course.

## Causal mediation estimation

Counterfactual-based direct and indirect effects can be estimated from
regression models, provided that the no confounding assumption hold. To
be able to do so, we need a model for the mediator and a model for the
outcome.

Model for the mediator:

$E(M|A = a, W = w) = \beta_0 + \beta_1a + \beta'_2w$

Model for the outcome:

$E(Y|A = a, M = m, W = w) = \sigma_0 + \sigma_1a + \sigma_2m + \sigma_3am + \sigma'_4w$

From these two regression models we can estimate the CDE, NDE and NIE.

## Example

The example here include a continuous treatment, mediator and outcome.
We use these equations on an additive scale because it is easier to
explain the intuition behind the estimands. Similar equations exist for
other types and combinations of variables.

```{r, echo = FALSE}
library(tidyverse)
```

```{r}
# This includes readr, or use library(readr) specifically
nhanes <- read_csv(here::here("data/NHANES.csv")) 
# keep the variables used in the analyses 
data <- nhanes %>% 
  dplyr::select(ID, 
                w1 = Age, 
                w2 = Gender, 
                w3 = Education, 
                w4 = Smoke100, 
                a = PhysActive,  # this is the exposure
                m = BMI,         # this is the mediator
                y = Pulse,       # this is the outcome
                y2 = Diabetes ) %>% 
  na.omit()
```

We can now run a model for the mediator only adjusting for a single
variable. In practice you would adjust for many more variables to
satisfy the confounding assumptions.

```{r}
lm_m <- lm(m ~ a + w1 , data = data)
summary(lm_m)
```

Here, we see that physical activity (a) is associated with lower pulse
(y).

Now, we run the model for the outcome, notice we have included an
interaction term.

```{r}
lm_y <- lm(y ~ a + m + a:m + w1 , data = data)
summary(lm_y)
```

In this model we can see:

-   higher physical activity (a) decreases pulse, independent of the
    mediator (BMI)

-   excess body weight (BMI) is associated with higher pulse,
    independent of physical activity

-   there is statistically significant interaction between physical
    activity (a) and BMI (m)

    Based on the coefficients of these two models, we can now estimate
    the different effects.

The equations are from @valeri_mediation_2013 and
@vanderweele_unification_2014.

### Controlled direct effect

The controlled direct effect can be obtained using this formula based on
the regression coefficients:

$CDE(m) = (\sigma_1 + \sigma_3*m)(a - a^*)$

$a^*$ is for a change from level $a^*$ (=control) to level $a$
(=intervention).

For the controlled direct effect we set m to a specific value.

Let's look at the distribution of m (BMI).

```{r}
data %>%
  select(m) %>%
  summary()

median_m <- round(median(data$m), digits = 1)
```

In this example we will set it to the median of BMI, at `r median_m`
which is a low level. You can also try to set BMI to 25kg/m2.

```{r}
CDE_m <- (lm_y$coefficients[2] + lm_y$coefficients[5] * median_m) * (1 - 0)
round(as.numeric(CDE_m), digits = 2)
```

The CDE is how much the outcome, here the pulse, would change on average
if the mediator, here BMI, were fixed at level (m = `r median_m`)
uniformly in the population but the treatment, physical activity, was
changed from 0 to 1.

The CDE answer the question, *what would be the effect of A on Y, when
fixing M at a specific value for everyone in the population*
@wang_g-computation_2015?

### (Pure) natural direct effect

PNDE can be obtained using this formula:

$PNDE = (\sigma_1 + \sigma_3 * (\beta_0 + \beta_1*a^* + \beta'_2*w))(a - a^*)$

We set the value for the confounder w1 for the interaction to be the
mean value of w1.

```{r}
mean_w1 <- mean(data$w1)
mean_w1
```

Then we calculate the PNDE.

```{r}
PNDE <- (lm_y$coefficients[2] + (lm_y$coefficients[5] * (lm_m$coefficients[1] + lm_m$coefficients[2] * 0 + lm_m$coefficients[3] * mean_w1))) * (1 - 0)
round(as.numeric(PNDE), digits = 2)
```

The PNDE is how much the outcome, pulse, would change if the treatment
a, physical activity, was set at 1 versus 0 but for each individual the
mediator was kept at the level it would have taken, for that individual,
in the absence of the exposure.

how pulse rate would change if the treatment (or exposure) physically
active (a=1) was set at 1 versus 0, while for each individual the
mediator (BMI) was kept at the level it would naturally be in the
**absense** of physical activity(**a=0**).

In other words, it measures the direct effect of physical activity on
pulse rate that doesn't work through changing BMI.

### (Total) natural direct effect

TNDE can be obtained by this formula:

$TNDE = (\sigma_1 + \sigma_3 * (\beta_0 + beta_1*a + \beta'_2*w))(a - a^*)$

```{r}
TNDE <- (lm_y$coefficients[2] + (lm_y$coefficients[5] * (lm_m$coefficients[1] + lm_m$coefficients[2] * 1 + lm_m$coefficients[3] * mean_w1))) * (1 - 0)

round(as.numeric(TNDE), digits = 2)
```

Here, the value of M is enabled to act (as opposed to the PNDE). Here we
see a little lower pulse rate, this means we're measuring the direct
effect of physical activity on pulse rate while allowing BMI to be at
the level it would naturally be with physical activity (**a=1**).

The TNDE asks the question: "to what extent does physical activity cause
lower pulse via pathways other than through BMI, allowing BMI to boost
up or tune down such effect at the same time?" @wang_g-computation_2015

### Total natural indirect effect

The TNIE can be obtained by this formula:

$TNIE = (\sigma_2 * \beta_1 + \sigma_3 * \beta_1 * a)(a - a^*)$

```{r}
TNIE <- ((lm_y$coefficients[3] * lm_m$coefficients[2]) + (lm_y$coefficients[5] * lm_m$coefficients[2] * 1)) * (1 - 0)

round(as.numeric(TNIE), digits = 2)
```

The TNIE is how much the outcome would change on average if the
treatment was fixed at level a = 1 but the mediator was changed from the
level it would take if a\* = 0 (physical inactive) to the level it would
take if a =1 (physical active).

Note that exposure has to have an effect on M otherwise this will be
zero.

The TNIE asks the question: to what extent does physical activity cause
pulse via BMI and the possible interaction between physical activity and
BMI in affecting pulse rate? In other words, the effect of exposure that
'would be prevented if the exposure did not cause the mediator' (i.e.,
the portion of the effect for which mediation is 'necessary')
@wang_g-computation_2015.

This is often the effect we are interested in in biomedical research for
questions regarding mediation.

### Pure natural indirect effect

The PNIE can be obtained by this formula:

$PNIE = (\sigma_2 * \beta_1 + \sigma_3 * \beta_1 * a^*)(a - a^*)$

```{r}
PNIE <- ((lm_y$coefficients[3] * lm_m$coefficients[2]) + (lm_y$coefficients[5] * lm_m$coefficients[2] * 0)) * (1 - 0)

round(as.numeric(PNIE), digits = 2)
```

The PNIE is different from the TNIE because it does not include the
interaction effect. We estimate the effect of red meat on inflammation
and then subsequent effect of inflammation on blood glucose.

The PNIE answer the question: "to what extent does physical activity
cause lower pulse rate via BMI only (i.e., due to physical activity
affecting BMI and subsequently, BMI affecting pulse rate), not
accounting for the possible interaction between physical activity and
BMI? In other words, the effect that the exposure would have had if 'its
only action were to cause the mediator' (i.e., the portion of the effect
for which mediation is ''sufficient'')" @wang_g-computation_2015.

### Total effect

The total effect can be decomposed as:

$TE = PNDE + TNIE$

```{r}
TE <- PNDE + TNIE

round(as.numeric(TE), digits = 2)
```

This is the overall effect of red meat on blood glucose levels. Higher
red meat is associated with higher blood glucose.

### Proportion mediation

From this, we can calculate the proportion mediated.

$PM = \frac{TNIE}{TE}$

```{r}
PM <- TNIE / TE

as.numeric(PM) * 100
```

21% of the association between physical activity and pulse rate is
mediated by the mediator BMI.

## Using R package

We can easily conduct mediation analysis using R package once you
understand the basic concepts of mediation.

```{r}

library(CMAverse)
res_rb <- cmest(
  data = data, 
  model = "rb",     #rb means regression based 
  outcome = "y",    #this is the name of outcome
  exposure = "a",   #this is the name of exposure
  mediator = "m",   #this is the name of mediator
  basec = c("w1"),  #this is confounding factors, you can add more confounders here 
  EMint = TRUE,     #whether include interaction, TRUE indicates to include 
  mreg = list("linear"), #specify the mediator model, we use linear regression here 
  yreg = "linear",  #specify the outcome model, we use linear regression here 
  astar = 0,       # the level of exposure (not exposed)
  a = 1,           # the level of exposure (exposed)
  mval = list(27.8),  #specifiy the value of controlled mediator 
  estimation = "paramfunc", #parameteric function 
  inference = "delta"
)
summary(res_rb)
```

## Defining estimands

Imagine we have a hypothetical randomized controlled trial where we give
participants treatment or no treatment on a specific outcome Y.

$Y^{a=1} - Y^{a=0}$

For mediation, we are also interested in the effect of a mediator on
this pathway. Now image that we also intervene on the mediator in a new
hypothetical randomized controlled trial.

$Y^{m=1} - Y^{m=0}$

Now consider if we, in the same trial, could intervene on both because
we are interested in whether treatment *causes* the outcome because it
*causes* the mediator.

::: callout-note
## Notation

$Y^a$ = a subject's outcome if treatment A were set, possible contrary
to fact, to a

$M^a$ = a subject's value of the mediator if the exposure A were set to
the value of a

$Y^{a,m}$ = a subject's outcome if A were set to a and M were set to m

$Y^{a,M_a}$ = a subject's outcome if A were set to a and M were set the
value m would have had had a been set to a. Note, this is a nested
counterfactual
:::

We can now define these estimands:

-   the controlled direct effect (CDE)
-   natural direct effect (NDE)
-   natural indirect effect (NIE)

### Controlled direct effect

The effect of A on Y not mediated through M. Fixing the value of M to m.

$Y^{a=1,m}$ - $Y^{a=0,m}$

We intervene on $a$ but fix $m$ to a certain value. The CDE is how much
the outcome would change on average if the mediator were fixed at level
m uniformly in the population but the treatment were changed from 0 to
1.

This could be relevant in the context of a change in a policy that
impacted the mediator for everyone. For instance, if air pollution was a
mediator between physical activity and cardiovascular disease risk. If a
new policy would change the level of air pollution for all while we
implement an intervention to increase biking in the city.

This effect is not used that often. But can be highly relevant in some
situations.

### (Pure) Natural direct effect

The effect that would remain, if we were to disable the pathway from
exposure to mediator.

$Y^{a=1,M_a=0}$ - $Y^{a=0,M_a=0}$

The PNDE is how much the outcome would change if the exposure was set at
a = 1 versus a\* = 0 but for each individual the mediator was kept at
the level it would have taken, for that individual, in the absence of
the exposure.

Note that the word "natural" refers to the nested counterfactual, the
level the mediator would have taken in the absence of exposure. What it
would naturally have been in the absence of exposure.

### Total natural direct effect

$Y^{a=1,M_a=1}$ - $Y^{a=0,M_a=1}$

Note, different from above in that the mediator is kept at the level it
would have taken in the **presence** of the exposure.

### (Total) Natural indirect effect

The effect of the mediator pathway.

$Y^{a=1,M_a=1}$ - $Y^{a=1,M_a=0}$

The NIE is how much the outcome would change on average if the exposure
were fixed at level a = 1 but the mediator were changed from the level
it would take if a\* = 0 to the level it would take if a = 1.

Note that exposure has to have an effect on M otherwise this will be
zero.

### Pure natural indirect effect

$Y^{a=0,M_a=1}$ - $Y^{a=0,M_a=0}$

Note, this is different from the TNIE in that the exposure is set to no
intervention.

### Interaction effects

More information on the interaction effects can be found in
@vanderweele_unification_2014.

Reference interaction:

$INT_{ref} = PNDE - CDE$

Mediation interaction:

$INT_{med} = TNIE - PNIE$

### Effect decomposition

Using the causal inference framework also allow for effect
decomposition, even when there are interaction and non-linearity.

Effect decomposition is important when we want to assess relative
contributions such as the proportion mediated and eliminated.

TE = PNDE + TNIE = TNDE + PNIE

More information about decomposition of effects can be found in
@vanderweele_unification_2014.

### Proportions

Proportion mediated:

$PM = NIE / TE$

Proportion eliminated:

$PE = (TE-CDE(m)) / TE$

## References
