# Introduction to causal mediation analysis

```{r setup}
#| include: false
#| cache: false
library(here)
library(DiagrammeR)
library(tidyverse)
```

In this session we will go over fundamental concepts of causal inference
in general and then focus more specifically on causal mediation
analysis.

::: callout-note
## Objective of this session

-   Foundational concepts in causal inference
-   From causal effects to causal mediation
-   Core causal assumptions
-   Causal mediation effects
-   Practical implementation using R
:::

### Recall the major limitations of traditional mediation methods:

-   Non-linearity
-   Interactions
-   Confounding
-   Multiple mediators

### Causal mediation analysis is an extension of the traditional approach:

-   Outlining all confounding assumptions needed
-   Handling non-linearity and interaction
-   Clearly defining estimands of interest
-   Different effect decompositions

## What is a cause?

We employ causal inference constantly in our daily lives. Consider a
simple example: when you press a light switch, you expect the light to
turn on. Or in another scenario, you anticipate that performing action A
will result in outcome Y. In some instances, we can identify
deterministic relationships where action A will lead to outcome Y. While
these straightforward cause-and-effect relationships are intuitive and
help us navigate the world, they become significantly more complex when
we explore health outcomes.

### Individual causal effect

When investigating health outcomes, we would ideally want to know if you
**do A**, then Y will happen. We could have a specific question:

> Will take the magic pill reduce my risk of developing T2D in 5 years?

To answer this question we would ideally have you take the magic pill
over 5 years and diagnose whether you get T2D. Then, we would turn back
time, and make you do not take the pill and then diagnose T2D. If there
is a difference between your two outcomes, then we say there is a causal
effect.

But we can never do this in the real world.

::: callout-note
## Notation

A = received treatment/intervention/exposure (e.g., 1 = intervention, 0
= no intervention)

Y = observed outcome (e.g., 1 = developed the outcome, 0 = no outcome)

$Y^{a=1}$ = Counterfactual outcome under treatment a=1 (i.e., the
outcome had everyone, counter to the fact, received treatment a = 1)

$Y^{a=0}$ = Counterfactual outcome under treatment a =0 (i.e., the
outcome had everyone, counter to the fact, received treatment a = 0)
:::

### Average causal effect

Instead, we can perform a randomized controlled trial. We now ask a
slightly different question:

> Will taking the magic pill reduce the risk of T2D?

We can randomly assigning one group to take the pill, and the other
group not to take the pill over 5 years. Then we compare the average
risk of T2D after 5 year in each of the groups. If there is a
difference, we could say there is an average causal effect
@hernan_definition_2004.

More formally we can now define a causal effect:

$E[Y^{a=1} = 1]-E[Y^{a=0} = 1]$

::: callout-note
## Notation

We now modify the terms a little.

$E[Y^{a=1}]$ the average counterfactual outcome, had all subjects in the
population received treatment a = 1.

$Pr[Y^{a=1}]$ the proportion of subjects that would have developed the
outcome Y had all subjects in the population of interest received
treatment a = 1.

$E[Y^{a=0}]$ the average counterfactual outcome, had all subjects in the
population received no treatment a = 0.

$Pr[Y^{a=0}]$ the proportion of subjects that would have developed the
outcome Y had all subjects in the population of interest received no
treatment a = 0.
:::

## From Causal effects to Causal mediation

When we want to understand not just **if**, but **how** the magic pill
affects risk of T2D, we need to consider mediating factors. For example:

```{r}
#| echo: false
grViz("
digraph {
  rankdir = LR;
  node [shape = box];
  
  Pill [label = 'Pill'];
  BMI [label = 'BMI'];
  T2D [label = 'T2D'];

  Pill -> T2D [label = 'Direct Effect'];
  Pill -> BMI [label = 'a'];
  BMI -> T2D [label = 'b'];
}
")
```

Will taking the pill reduce my BMI, which in turn decreases the risk of
T2D? This requires us to decompose the total effect into direct and
indirect pathways.

![cross-world
counterfacturals](/images/decompose1.png){#fig-decompose1}

For mediation effects, we need to consider scenarios that cross between
different possible worlds, such as: What would my risk of getting T2D be
if I take the pill (treatment = 1), BUT my BMI remained at the level it
would have been had I NOT taken the pill (mediator at level from
treatment = 0), compared to the risk of T2D if I had taken the pill, and
my BMI were at the level it would naturally reach under treatment
(treatment =1).

These **cross-world counterfactuals** combine elements from two
incompatible realities. They cannot be directly observed even in
principle, which creates fundamental identification challenges.

## Causal Mediation Estimation Approaches

Counterfactual-based direct and indirect effects can be estimated from
regression models, provided that the no confounding assumption hold. To
be able to do so, we need a model for the mediator and a model for the
outcome.

::: callout-note
Causal approaches define effects using the counterfactural framework,
before estimating th effects using statistical methods.
:::

Model for the mediator:

$E(M|A = a, W = w) = \beta_0 + \beta_1a + \beta'_2w$

Model for the outcome:

$E(Y|A = a, M = m, W = w) = \sigma_0 + \sigma_1a + \sigma_2m + \sigma_3am + \sigma'_4w$

From these two regression models we can estimate the direct and indirect
effect.

Direct effect: $E[(Y(a=1,M_a=0))$ - $Y(a=0,M_a=0))]$

Indirect effect: $E[(Y(a=1,M_a=1))$ - $Y(a=1,M_a=0))]$

While regression-based approaches are commonly used, several alternative
estimation methods are available, including Inverse probability
weighting (IPW), G-computation, and structural equation models (SEM).

Note: additional approaches also exists, but we will not focus on these
in this course.

## Casual Mediation Effects

### Defining estimands

Imagine we have a hypothetical randomized controlled trial where we give
participants treatment or no treatment on a specific outcome Y.

$Y^{a=1} - Y^{a=0}$

For mediation, we are also interested in the effect of a mediator on
this pathway. Now image that we also intervene on the mediator in a new
hypothetical randomized controlled trial.

$Y^{m=1} - Y^{m=0}$

Now consider if we, in the same trial, could intervene on both because
we are interested in whether treatment *causes* the outcome because it
*causes* the mediator.

::: callout-note
## Notation

$Y^a$ = a subject's outcome if treatment A were set, possible contrary
to fact, to a

$M^a$ = a subject's value of the mediator if the exposure A were set to
the value of a

$Y^{a,m}$ = a subject's outcome if A were set to a and M were set to m

$Y^{a,M_a}$ = a subject's outcome if A were set to a and M were set the
value m would have had had a been set to a. Note, this is a nested
counterfactual
:::

We can now define these estimands:

-   the controlled direct effect (CDE)
-   natural direct effect (NDE)
-   natural indirect effect (NIE)

#### Controlled direct effect

$Y^{a=1,m}$ - $Y^{a=0,m}$

We intervene on $a$ but fix $m$ to a certain value. The CDE is how much
the outcome would change on average if the mediator were fixed at level
m uniformly in the population but the treatment were changed from 0 to
1.

This could be relevant in the context of a change in a policy that
impacted the mediator for everyone. For instance, if air pollution was a
mediator between physical activity and cardiovaular disease risk. If a
new policy would change the level of air pollution for all while we
implement an intervention to increase biking in the city.

This effect is not used that often. But can be highly relevant in some
situations.

![controled direct
effect](/images/cde.png){#fig-cde}

#### (Pure) Natural direct effect

$Y^{a=1,M_a=0}$ - $Y^{a=0,M_a=0}$

![pure natural direct
effect](/images/pnde.png){#fig-pnde}

The PNDE is how much the outcome would change if the exposure was set at
a = 1 versus a\* = 0 but for each individual the mediator was kept at
the level it would have taken, for that individual, in the absence of
the exposure.

Note that the word "natural" refers to the nested counterfactual, the
level the mediator would have taken in the absence of exposure.

#### (Total) Natural indirect effect

$Y^{a=1,M_a=1}$ - $Y^{a=1,M_a=0}$

![total natural indirect
effect](/images/tnie.png){#fig-tnie}

The NIE is how much the outcome would change on average if the exposure
were fixed at level a = 1 but the mediator were changed from the level
it would take if a\* = 0 to the level it would take if a = 1.

Note that exposure has to have an effect on M otherwise this will be
zero.

::: callout-note
## Decomposition

TE = PNDE + TNIE this decomposition usually is for research question:
'is there a mediated effect?' or 'is the causal effect mediated by the
proposed mediator?' @nguyen2021clarifying
:::

::: callout-note
The other way to decompose the TE is by Total natural direct effect and
pure natural indirect effect.

#### Total natural direct effect

TNDE = $Y^{a=1,M_a=1}$ - $Y^{a=0,M_a=1}$

Note, different from above in that the mediator is kept at the level it
would have taken in the **presence** of the exposure.

#### Pure natural indirect effect

PNIE = $Y^{a=0,M_a=1}$ - $Y^{a=0,M_a=0}$

Note, this is different from the TNIE in that the exposure is set to no
intervention.

TE = PNIE + TNDE this decomposition usually is for research question:
'does the exposure influence the outcome in other ways, not through this
mediator?' @nguyen2021clarifying
:::

### Effect decomposition

Using the causal inference framework also allow for effect
decomposition, even when there are interaction and non-linearity.

Effect decomposition is important when we want to assess relative
contributions such as the proportion mediated and eliminated.

TE = PNDE + TNIE = TNDE + PNIE

More information about decomposition of effects can be found in
@vanderweele_unification_2014.

### Proportions

Proportion mediated:

$PM = NIE / TE$

Proportion eliminated:

$PE = (TE-CDE) / TE$

## Why Causal Mediation Analysis?

Neither the product method nor the difference method in traditional
mediation analysis can effectively account for interaction effects or
non-linear relationships between variables. Traditional methods also
lack attention to confounding and temporallity.

Causal mediation analysis addresses these limitations through more
sophisticated approaches. Specifically, it can:

-   Constructing a comprehensive causal model that explicitly
    incorporates non-linear relationships and interaction effects.
    Counterfactual approaches can better handle complex, non-parametric
    relationships.

-   Decomposition of effects: Counterfactual methods allow for more
    nuanced decomposition of total effects into direct and indirect
    effects, especially in the presence of treatment-mediator
    interactions.

-   Counterfactual frameworks extend more naturally to settings with
    multiple mediators

## Core Causal Assumptions

### Association vs Causation

What makes it complicated to estimate a causal effect is that we cannot
observe the outcome under different treatments.

When we only have a subset of the outcomes, we have an association. This
is illustrated in @fig-causation-association.

![Association versus Causation: When we observe only a subset of the
outcomes across exposure groups, we have an association rather than
causation.](/images/causation-association.png){#fig-causation-association}

::: callout-note
-   Causal Assumptions

    -   Exchangeability

    The risk of the outcome in A = 1 would have been the same as the
    risk of the outcome in A = 0, had those in A = 1 received A = 0.

    Think about a randomized trial where you, by mistake, give the
    intervention to the other group. The effect should be the same as
    the one you would have observed had the groups been correct.

    We can also have conditional exchangeability. The risk is similar in
    subsets of the population. That could be within the levels of a
    variable W (e.g., education).

    -   Consistency

    The treatments under comparison are well-defined and correspond to
    the versions of treatment observed in the data:

    -   Precise definition of $Y^a$ via a
    -   Link counterfactual outcomes with observed outcomes

    $Y^a = Y$ for every individual with A = a.

    The observed outcome for all treated equals the outcome if they had
    received the treatment (the potential outcome).

    -   Positivity

    The probability of receiving every value of treatment conditional on
    L is greater than zero. In other words, there must be a probability
    of being assigned to each treatment level.

    Image if we don't have anyone with A = 0 among those with L=1 (e.g.
    long education), then we cannot estimate the conditional probability
    of the outcome.
:::

### Confounding assumptions in Causal Mediation Analysis

Confounding assumptions are critical for valid causal inferences in
mediation analysis. These assumptions specify the conditions under which
we can interpret our estimates as truly causal rather than merely
associational.

<div>

1.  no unmeasured exposure-outcome confounding

2.  no unmeasured exposure-mediator confounding

3.  no unmeasured mediator-outcome confounding

4.  mediator-outcome confounding not affected by the exposure/treatment
    (post-treatment confounder or intermediate confounder)

</div>

By using DAGs, the assumptions about confounding are made much more
explicit.

```{r}
#| echo: false

grViz("
digraph {
  graph []
  node [shape = plaintext]
    U [label = 'U']
    A [label = 'A']
    M [label = 'M']
    Y [label = 'Y']
  edge [minlen = 2]
    A->M
    M->Y
    A->Y
    U->A
    U->Y
    U->M
{rank = same; U; A; M; Y; }
}
")
```

We can see that we not only have to take confounding between the
exposure and outcome into account, but we also have to take
mediator-outcome ($A \leftarrow W \rightarrow Y$) confounding into
account.

In addition, we can have a more complicated situation where the exposure
also impacts another mediator that is also a mediator-outcome
confounder.

Below DAG shows where the mediator-outcome relation is affected by
treatment(or exposure):

```{r}
#| echo: false

grViz("
digraph {
  graph []
  node [shape = plaintext]
    U1 [label = 'U1']
    A [label = 'A']
    M [label = 'M']
    Y [label = 'Y']
    U2 [label = 'U2']
  edge [minlen = 2]
    A->M
    M->Y
    A->Y
    U1->A
    U1->Y
    U1->M
    U2->M
    U2->Y
    A->U2
    U1->U2
{rank = same; U1; A; M; Y; }
{rank = max; U2 ; }
}
")
```

From the DAG rules, we have a special problem that we cannot solve with
traditional regression approaches. If we adjust for U2 we open a
backdoor path by adjusting for the collider
$U1 \rightarrow U2 \leftarrow A$.

## Practical Examples using R package

We can easily conduct mediation analysis using R package once you
understand the basic concepts of mediation.

::: callout-note
## Before you run the analytical models

-   make a DAG display the assumed causal model
-   check the confounding assumptions for each path
-   ensure the temporal ordering of the variables
:::

```{r}
# This includes readr, or use library(readr) specifically
nhanes <- read_csv(here::here("data/NHANES.csv"))
data <- nhanes %>%
  dplyr::select(ID,
    w1 = Age,
    w2 = Gender,
    w3 = Education,
    w4 = Smoke100,
    a = PhysActive,
    m = BMI,
    y = Pulse,
    y2 = Diabetes
  ) %>%
  na.omit()
```

First, load the library:

```{r}
library(CMAverse)
```

Next, we can plot the DAG based on the research question:

```{r}
cmdag(
  outcome = "y", exposure = "a", mediator = c("m"),
  basec = c("W1", "W2", "W3", "W4"), postc = NULL, node = FALSE, text_col = "black"
)
```

Then we can start estimate causal mediation effects using the function
'cmest'.

```{r}
res_rb <- cmest(
  data = data,
  model = "rb", # rb means regression based
  outcome = "y", # this is the name of outcome
  exposure = "a", # this is the name of exposure
  mediator = "m", # this is the name of mediator
  basec = c("w1"), # this is confounding factors, you can add more confounders here
  EMint = TRUE, # whether include interaction, TRUE indicates to include
  mreg = list("linear"), # specify the mediator model, we use linear regression here
  yreg = "linear", # specify the outcome model, we use linear regression here
  astar = 0, # the level of exposure (not exposed)
  a = 1, # the level of exposure (exposed)
  mval = list(27.8), # specifiy the value of controlled mediator
  estimation = "paramfunc", # parameteric function
  inference = "delta"
)
summary(res_rb)
```

```{r}
sim_rb <- cmest(
  data = data,
  model = "rb", # rb means regression based
  outcome = "y", # this is the name of outcome
  exposure = "a", # this is the name of exposure
  mediator = "m", # this is the name of mediator
  basec = c("w1"), # this is confounding factors, you can add more confounders here
  EMint = TRUE, # whether include interaction, TRUE indicates to include
  mreg = list("linear"), # specify the mediator model, we use linear regression here
  yreg = "linear", # specify the outcome model, we use linear regression here
  astar = 0, # the level of exposure (not exposed)
  a = 1, # the level of exposure (exposed)
  mval = list(27.8), # specifiy the value of controlled mediator
  estimation = "imputation", ## use imputation
  inference = "bootstrap",
  nboot = 20
)
summary(sim_rb)
```

```{r}
ggcmest(res_rb) +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 30, vjust = 0.8))
```

## Example-not using R package

You can also estimate causal mediation effects using regression-based
models by implementing the calculation steps manually

```{r, echo = FALSE}
library(tidyverse)
```

We can now run a model for the mediator only adjusting for a single
variable. In practice you would adjust for many more variables to
satisfy the confounding assumptions.

```{r}
lm_m <- lm(m ~ a + w1, data = data)
summary(lm_m)
```

Here, we see that physical activity (a) is associated with lower BMI(m).

Now, we run the model for the outcome, notice we have included an
interaction term.

```{r}
lm_y <- lm(y ~ a + m + a:m + w1, data = data)
summary(lm_y)
```

In this model we can see:

-   higher physical activity (a) decreases pulse, independent of the
    mediator (BMI)

-   excess body weight (BMI) is associated with higher pulse,
    independent of physical activity

-   there is statistically significant interaction between physical
    activity (a) and BMI (m)

    Based on the coefficients of these two models, we can now estimate
    the different effects.

The equations are from @valeri_mediation_2013 and
@vanderweele_unification_2014.

### Controlled direct effect

The controlled direct effect can be obtained using this formula based on
the regression coefficients:

$CDE(m) = (\sigma_1 + \sigma_3*m)(a - a^*)$

$a^*$ is for a change from level $a^*$ (=control) to level $a$
(=intervention).

For the controlled direct effect we set m to a specific value.

Let's look at the distribution of m (BMI).

```{r}
data %>%
  select(m) %>%
  summary()
median_m <- round(median(data$m), digits = 1)
```

In this example we will set it to the median of BMI, at `r median_m`
which is a low level. You can also try to set BMI to 25kg/m2.

```{r}
CDE_m <- (lm_y$coefficients[2] + lm_y$coefficients[5] * median_m) * (1 - 0)
round(as.numeric(CDE_m), digits = 2)
```

The CDE is how much the outcome, here the pulse, would change on average
if the mediator, here BMI, were fixed at level (m = `r median_m`)
uniformly in the population but the exposure-physical activity, was
changed from 0 to 1.

The CDE answer the question, *what would be the effect of A on Y, when
fixing M at a specific value for everyone in the population*
@wang_g-computation_2015?

### (Pure) natural direct effect

PNDE can be obtained using this formula:

$PNDE = (\sigma_1 + \sigma_3 * (\beta_0 + \beta_1*a^* + \beta'_2*w))(a - a^*)$

We set the value for the confounder w1 for the interaction to be the
mean value of w1.

```{r}
mean_w1 <- mean(data$w1)
mean_w1
```

Then we calculate the PNDE.

```{r}
PNDE <- (lm_y$coefficients[2] + (lm_y$coefficients[5] * (lm_m$coefficients[1] + lm_m$coefficients[2] * 0 + lm_m$coefficients[3] * mean_w1))) * (1 - 0)
round(as.numeric(PNDE), digits = 2)
```

The PNDE is how much the outcome, pulse, would change if the
exposure-physical activity, was set at 1 versus 0 but for each
individual the mediator was kept at the level it would have taken, for
that individual, in the absence of the exposure.

how pulse rate would change if the exposure-physically active (a=1) was
set at 1 versus 0, while for each individual the mediator (BMI) was kept
at the level it would naturally be in the **abcense** of physical
activity(**a=0**).

In other words, it measures the direct effect of physical activity on
pulse rate that doesn't work through changing BMI.

### (Total) natural direct effect

TNDE can be obtained by this formula:

$TNDE = (\sigma_1 + \sigma_3 * (\beta_0 + beta_1*a + \beta'_2*w))(a - a^*)$

```{r}
TNDE <- (lm_y$coefficients[2] + (lm_y$coefficients[5] * (lm_m$coefficients[1] + lm_m$coefficients[2] * 1 + lm_m$coefficients[3] * mean_w1))) * (1 - 0)
round(as.numeric(TNDE), digits = 2)
```

Here, the value of M is enabled to act (as opposed to the PNDE). Here we
see a little lower pulse rate, this means we're measuring the direct
effect of physical activity on pulse rate while allowing BMI to be at
the level it would naturally be with physical activity (**a=1**).

The TNDE asks the question: "to what extent does physical activity cause
lower pulse via pathways other than through BMI, allowing BMI to boost
up or tune down such effect at the same time?" @wang_g-computation_2015

### Total natural indirect effect

The TNIE can be obtained by this formula:

$TNIE = (\sigma_2 * \beta_1 + \sigma_3 * \beta_1 * a)(a - a^*)$

```{r}
TNIE <- ((lm_y$coefficients[3] * lm_m$coefficients[2]) + (lm_y$coefficients[5] * lm_m$coefficients[2] * 1)) * (1 - 0)
round(as.numeric(TNIE), digits = 2)
```

The TNIE is how much the outcome would change on average if the exposure
was fixed at level a = 1 but the mediator was changed from the level it
would take if a\* = 0 (physical inactive) to the level it would take if
a =1 (physical active).

Note that exposure has to have an effect on M otherwise this will be
zero.

The TNIE asks the question: to what extent does physical activity cause
pulse via BMI and the possible interaction between physical activity and
BMI in affecting pulse rate? In other words, the effect of exposure that
'would be prevented if the exposure did not cause the mediator' (i.e.,
the portion of the effect for which mediation is 'necessary')
@wang_g-computation_2015.

This is often the effect we are interested in in biomedical research for
questions regarding mediation.

### Pure natural indirect effect

The PNIE can be obtained by this formula:

$PNIE = (\sigma_2 * \beta_1 + \sigma_3 * \beta_1 * a^*)(a - a^*)$

```{r}
PNIE <- ((lm_y$coefficients[3] * lm_m$coefficients[2]) + (lm_y$coefficients[5] * lm_m$coefficients[2] * 0)) * (1 - 0)
round(as.numeric(PNIE), digits = 2)
```

The PNIE is different from the TNIE because it does not include the
interaction effect. We estimate the effect of physical activity on BMI
and then subsequent effect of BMI on pulse rate.

The PNIE answer the question: "to what extent does physical activity
cause lower pulse rate via BMI only (i.e., due to physical activity
affecting BMI and subsequently, BMI affecting pulse rate), not
accounting for the possible interaction between physical activity and
BMI? In other words, the effect that the exposure would have had if 'its
only action were to cause the mediator' (i.e., the portion of the effect
for which mediation is ''sufficient'')" @wang_g-computation_2015.

### Total effect

The total effect can be decomposed as:

$TE = PNDE + TNIE$

```{r}
TE <- PNDE + TNIE
round(as.numeric(TE), digits = 2)
```

This is the overall effect of physical activity on pulse rate.

### Proportion mediation

From this, we can calculate the proportion mediated.

$PM = \frac{TNIE}{TE}$

```{r}
PM <- TNIE / TE
as.numeric(PM) * 100
```

21% of the association between physical activity and pulse rate is
mediated by the mediator BMI.

## References
