---
editor: 
  markdown: 
    wrap: 72
---

# Motivation

## Mediation analysis in Epidemiology

```{r setup}
#| include: false 
library(DiagrammeR)
```

::: callout-tip
Mediation analysis is the *study of pathways and mechanisms through
which an exposure or intervention impacts an outcome*.
:::

In clinical and epidemiological research, the primary focus is often on
determining whether a specific exposure or intervention has an effect on
a disease or health outcome (aetiological epidemiology). Once this
effect is established, the next natural question is to explore the
"black box"—the underlying mechanisms that explain how the exposure (or
intervention) leads to the observed outcome. As we are not only
interested in whether an exposure/intervention has an effect, but in how
this effect works @vanderweele_mediation_2016.

Mediation analysis helps us to define an answer to this HOW question, by
quantitatively exploring the pathways and mechanisms that explain the
causal relationship between the exposure (or intervention or treatment)
and the outcome. By doing so, mediation analysis helps to open the black
box.

## Motivation for mediation analysis

1.  **Explanation and Understanding**:

    Unpacking how and why certain factors (e.g., physical activity)
    impact outcomes (e.g., diabetes risk) through intermediate variables
    (e.g., weight loss).

    *Example*: Does body mass index mediate the relationship between
    diet and type 2 diabetes in mid-life?

2.  **Confirmation and Refutation of Theory**:

    Testing theories about the mechanisms driving health outcomes and
    determining whether certain factors mediate the relationships
    between exposures and outcomes.

    *Example*: Is the relationship between lower socioeconomic status
    and cardiovascular disease explained (mediated) through high lipid
    levels?

3.  **Refining Interventions**:

    Mediation analysis may help us to improve the effectiveness of
    interventions by allowing us to target key mediators.

    *Example*: Let's consider the evidence from the landmark diabetes
    prevention trials: The Diabetes Prevention Programme
    @knowler_reduction_2002, The Diabetes Prevention Study
    @tuomilehto_prevention_2001, and The Da Qin study @pan_effects_1997.
    These studies randomized high diabetes risk individuals to either a
    lifestyle intervention (consisting of diet modfication in
    combination with physical activity) or routine health advice and
    care. The main finding was a \~58% lower risk of incidence type 2
    diabetes in the lifestyle intervention group compared to the routine
    care group.

```{r, echo=FALSE}
graph <- grViz("
digraph {
  graph [rankdir = LR]  # Left to right graph
  node [shape = plaintext] 
  X [label = 'Lifestyle intervention'] 
  M [label = 'Weight loss'] 
  Y [label = 'Type 2 diabetes']
  edge [minlen = 2]
  X -> M
  M -> Y
}
")
graph
```

We might be interested in further refining the intervention so as to
increase the magnitude of the effect. It would be interesting to know if
the lifestyle changes reduce the risk of diabetes only because they
lower BMI (effect mediated through BMI) or if lifestyle has a direct
impact on diabetes risk that does not require any change in BMI. This
may help us understand how important BMI targets are in future lifestyle
modification programs.

```{r,, echo=FALSE}
library(DiagrammeR)
grViz("
digraph {
  graph []
  node [shape = plaintext]
    X [label = 'Lifestyle intervention']
    M [label = 'Weight loss']
    N [label = '?']
  Y [label = 'Type 2 diabetes']
  edge [minlen = 2]
    X->M
    M->Y
    N->M
   {rank = same; M; X; Y; }
   {rank = min; N; }
}
")
```

::: callout-note
## Can you find a relevant mediation question in your own research?

Spend 3 minutes talking to your neighbour and discuss an mediation
question that may be relevant in your own research.
:::

## Motivations for running this workshop

The methodological framework for conducting **mediation analysis** has
evolved rapidly in recent years, making it a crucial tool for
understanding complex relationships in research. Initially rooted in
basic regression techniques, mediation analysis has advanced into
**causal mediation analysis**, enabling researchers to unravel the
intricate pathways between exposures, mediators, and outcomes.

This short workshop aims to introduce the **basic statistical
framework**, key assumptions, and practical applications of both
**traditional** and **modern** mediation analysis techniques. Through
**real-world data examples**, we’ll explore how mediation analysis can
enhance your research and understanding of causal relationships.

Despite its growing importance, **mediation analysis** is often used
incorrectly. A recent **systematic review** on the reporting quality of
studies applying mediation analysis revealed that many studies provide
incomplete reporting, which hinders both **reproducibility** and the
**translation** of findings into practice [@rrn_systematic_2022] .

Our goal is for you to learn how to apply **mediation analysis
correctly** and leverage it as a powerful tool to advance your research.
By improving the quality and rigor of mediation analysis in your work,
you can ensure more reliable findings and meaningful insights.

## References
