---
title: "Motivation"
format: html
editor: visual
---

# Motivation for mediation analysis

<div>

What is your motivation for this course on mediation analysis?

Spend 3 minutes talking to your neighbour and discuss your own motivations for participating in this course.

</div>

## Mediation analysis in Epidemiology 

```{r setup}
#| include: false 
library(DiagrammeR)
```

Mediation analysis is the study of pathways and mechanisms through which an exposure or intervention impacts an outcome. In clinical and epidemiological research, the primary focus is often on determining whether a specific intervention has an effect on a disease or health outcome. Once this effect is established, the next natural question is to explore the "black box"—the underlying mechanisms that explain how the intervention (or exposure) leads to the observed outcome. As we are not only interested in whether an intervention works, but how it works (\[\@vanderweele_mediation_2016\]).

Mediation analysis provides an answer to this **WHY** question, by uncovering the pathways and mechanisms that explain the causal relationship between the exposure (e.g., intervention or treatment) and the outcome. By doing so, mediation analysis helps to open the black box, showing not only if an intervention works but how it works, and guiding more effective and targeted interventions.

## Motivation for mediation analysis

1.  **Explanation and Understanding**: Unpacking how and why certain factors (e.g., physical activity) impact outcomes (e.g., diabetes risk) through intermediary variables (e.g., weight loss).

*Example*: Does body mass index mediate the relationship between diet and type 2 diabetes in mid-life?

2.  **Confirmation and Refutation of Theory**: Testing theories about the mechanisms behind health outcomes and determining whether certain factors mediate the relationships between exposures and outcomes.

*Example*: Is the relationship between lower socioeconomic status and cardiovascular disease explained (mediated) through high allostatic load?

2.  **Refining Interventions**: Using mediation analysis to improve the effectiveness of interventions by targeting key mediators, leading to better and more sustainable outcomes.

*Example*: Let's consider the evidence from the landmark prevention trials. The Diabetes Prevention Programme @knowler_reduction_2002, The Diabetes Prevention Study @tuomilehto_prevention_2001, and The Da Quin study @pan_effects_1997. These studies randomized high risk individuals to either metformin or a lifestyle intervention consisting of a physical activity or physical activity + diet. The main finding was a \~58% lower risk of incidence type 2 diabetes in the lifestyle intervention group compared to the metformin group.

```{r, echo=FALSE}
library(DiagrammeR)
grViz("
digraph {
  graph [rankdir = LR]  # Left to right graph
  node [shape = plaintext] 
  X [label = 'Lifestyle intervention'] 
  M [label = 'Weight loss'] 
  Y [label = 'Type 2 diabetes']
  edge [minlen = 2]
  X -> M
  M -> Y
}
")
```

We might be interested in further refining the intervention so as to increase the magnitude of the effect. This might be done by altering or improving components of the intervention that target a particular mechanism for the outcome and discarding those components of which perhaps are not relevant for the outcome

When implementing the same lifestyle interventions in real world settings, lifestyle interventions show smaller effect and the long-term adoption and sustainability over time were some of the challenges to the effectiveness of these prevention efforts

```{r, echo=FALSE}
library(DiagrammeR)
grViz("
digraph {
  graph []
  node [shape = plaintext]
    X [label = 'Lifestyle intervention']
    M [label = 'Weight loss']
    N [label = '?']
  Y [label = 'Type 2 diabetes']
  edge [minlen = 2]
    X->M
    M->Y
    N->M
   {rank = same; M; X; Y; }
   {rank = min; N; }
}
")
```

## Motivations for running this course

The methodological framework for conducting **mediation analysis** has evolved rapidly in recent years, making it a crucial tool for understanding complex relationships in research. Initially rooted in basic regression techniques, mediation analysis has advanced into **causal mediation analysis**, enabling researchers to unravel the intricate pathways between exposures, mediators, and outcomes.

This short course aims to introduce the **basic statistical framework**, key assumptions, and practical applications of both **traditional** and **modern** mediation analysis techniques. Through **real-world data examples**, we’ll explore how mediation analysis can enhance your research and understanding of causal relationships.

Despite its growing importance, **mediation analysis** is often used incorrectly. A recent **systematic review** on the reporting quality of studies applying mediation analysis revealed that many studies provide incomplete reporting, which hinders both **reproducibility** and the **translation** of findings into practice (\[\@rrn_systematic_2022\]).

Our goal is for you to learn how to apply **mediation analysis correctly** and leverage it as a powerful tool to advance your research. By improving the quality and rigor of mediation analysis in your work, you can ensure more reliable findings and meaningful insights.
