# Pre-course tasks {#sec-pre-course}

In order to
participate in this course actively, you recommend you complete the
pre-course tasks. 

These tasks are designed to make it easier for everyone to start the
course with everything ready to go.

Depending on your skills and knowledge, these tasks could take
between **1 to 2 hours to finish**.

If you have any issues, please write us an email or ask during the
registration on the morning of the first day of the course.

## List of tasks

Here's a quick overview of the tasks you need to do. Specific details
about them are found as you work through this section.

::: callout-note 
1. Install the necessary programs and the right version of the programs. 
2. Get familiar with Directed Acyclic Graphs. 
3. Download the data or the repository in @sec-download-data
4. Read the Code of Conduct
:::

## 1. Installing programs {#sec-intall-r-packages}

If you are a workshop attendee and have any trouble
with installation locally, please reach out to us directly!

1️⃣ Install R (Latest Version): If you have not already done so, please
download and install the latest version of R based on your operating
system:

<https://cran.r-project.org/>

2️⃣ Install
RStudio: This is a separate program that provides an intuitive user
interface with R. 

Visit RStudio Download Page:
<https://posit.co/products/open-source/rstudio/> Click "Download
RStudio" and follow the instructions.

3️⃣ Check Your R Version If you already have R
installed, check your version to ensure compatibility: \>\>\>\>\>\>\>
c0858882aefc0e9dc24c8e312fba09cea0bcba8d:Precourse_files/Precourse.qmd

```{r}
#| echo: true
#| eval: false
#| results: asis
R.Version()
```

If your R version is 4.3.1 or later, you're good to go!

4️⃣ Install Required R Packages: To ensure you have all necessary
packages, run the following command in R:

```{r}
#| echo: true
#| eval: false
install.packages("pak")
pak::pkg_install("tidyverse")
pak::pkg_install("DiagrammeR")
pak::pkg_install("downlit")
pak::pkg_install("furrr")
pak::pkg_install("here")
pak::pkg_install("knitr")
pak::pkg_install("mediation")
pak::pkg_install("NetCoupler")
pak::pkg_install("rmarkdown")
pak::pkg_install("survival")
pak::pkg_install("xml2")
pak::pkg_install("here")
pak::pkg_install("BS1125/CMAverse")
pak::pkg_install("ggdag")
pak::pkg_install("ggplot2")
```

If installation of a package does not work using the "pak" package, then
use:

```{r}
#| echo: true
#| eval: false
#| results: asis

install.packages("name-of-package")
```

If you have issues with the CMA package, you could also try this:

```{r}
#| echo: true
#| eval: false
#| results: asis

install.packages("devtools")
devtools::install_github("BS1125/CMAverse")
```

## 2. DAG in pre-course {#sec-code-of-conduct}

We designed an introduction to Directed Acyclic Graph,
please get yourself familiar with this tool.

link: [Introduction to DAGs](/preamble/Introduction-to-DAGs.qmd)

## 3. Download the data and repository {#sec-download-data}

You can follow all the teaching material on this website. For the
exercises, you will need to use the datasets we have created. You can
either download the entire repository (recommended) or simply download
the data from our github. We recommend that you do not do this before
March 31, in the afternoon as we will making updates until then.

### Download repository

1.  Open a new session of R in RStudio. If you are in a project, go out
    of that project.

2.  Using RStudio's New Project menu item: "File -\> New Project -\>
    Version control -\> Git"

3.  Keeping RStudio open, open your browser and go to our
    ??[repository](https://github.com/steno-aarhus/mediation-analysis-course)

4.  Then, Click on the green "code" icon and copy-paste

5.  Now go back to R and put in what you just copy-pasted under
    "Repository URL." You should also select where this project should.

6. Click "create project" and you should now have all materials there.

### Download the data

If you want to only download the course data. You can go to
[here](https://github.com/steno-aarhus/mediation-analysis-course/tree/main/data)
for the Framingham, NHANES and simulated data. Click "download raw file"
in the top right corner after you have clicked on the specific file.

## 4. Code of conduct

We are dedicated to providing a welcoming and supportive environment for
all people, regardless of background or identity. As such, we do not
tolerate behaviour that is disrespectful to our teachers or learners or
that excludes, intimidates, or causes discomfort to others. We do not
tolerate discrimination or harassment based on characteristics that
include, but are not limited to, gender identity and expression, sexual
orientation, disability, physical appearance, body size, citizenship,
nationality, ethnic or social origin, pregnancy, familial status,
veteran status, genetic information, religion or belief (or lack
thereof), membership of a national minority, property, age, education,
socio-economic status, technical choices, and experience level.

Everyone who
participates in this workshop’s activities is required to conform to
this Code of Conduct. It applies to all spaces associated with this
workshop including, but not limited to, group work, emails, and any
online forums such as GitHub and X(Twitter). By participating,
participants indicate their acceptance of the procedures by which the
instructors and organizers resolves any Code of Conduct incidents.

See you at the course!